---
title: "LRcellTypeMarkers: Marker gene information for LRcell R Bioconductor package using Bioconductor's ExperimentHub"
author: "Wenjing Ma"
date: "`r doc_date()`"
bibliography: library.bib
vignette: >
  %\VignetteIndexEntry{LRcellTypeMarkers: Marker gene information for LRcell R Bioconductor package using Bioconductor's ExperimentHub}
  %\VignetteEngine{knitr::rmarkdown}
output: 
  BiocStyle::html_document
---

# Download cell-type specific marker gene information

LRcellTypeMarkers currently provides gene enrichment scores for mouse whole brain
and human prefrontal cortex. These gene enrichment scores were calculated
using the algorithm proposed in [@marques2016oligodendrocyte]. The detailed 
procedures are described in `scripts/make-data.R`. Users are encouraged
to download the data, generate customized cell-type specific marker genes and
apply it in on bulk DE genes using [LRcell](https://github.com/marvinquiet/LRcell). 

The example below shows how to download the data using ExperimentHub.

```{r intro}
library(ExperimentHub)
eh <- ExperimentHub()
eh <- query(eh, "LRcellTypeMarkers")
eh ## show entries of LRcellTypeMarkers package
```

According to the description, users can choose the most suitable marker gene
information for further analysis. Here, we used mouse frontal cortex marker genes
as an example, which has the title as **EH4548**.

```{r extract_data}
mouse_FC <- eh[['EH4548']]
mouse_FC[1:6, 1:6]  # show head of the data
```

From the data, we can observe that the rows are gene symbols and the columns are 
the sub-cell types (or clusters). The values indicate the gene enrichment level 
in a sub-cell type (or cluster). Higher the values are, more unique the genes 
are in the certain cell types. Users can sort the enrichment score in a 
descending order and select top 100 genes for each sub-cell type (or cluster) 
using the following command or the function named `get_markergenes()` in LRcell.

Here, we take a glance at the first 6 clusters and the first 6 marker genes in 
each cluster.

```{r sort}
library(LRcell)
FC_marker_genes <- get_markergenes(mouse_FC, method="LR", topn=100)
head(lapply(FC_marker_genes, head))  # glance at the marker genes
```

**Note**: For MSigDB C8 datasets, the marker genes are only available for 
Logistic Regression (LR) analysis. 

# Usage of cell-type specific markers
LRcellTypeMarkers package is a data source for running LRcell analysis which 
identifies (sub-)cell types are transcriptionally enriched in bulk differentially
expressed genes (DEGs) experiments. Here, we demonstrate an example on using provided
mouse brain Frontal Cortex marker genes on Alzheimer's Disease (AD) to identify Microglia.

First, we load the sample bulk DEGs vector provided by LRcell package which 
contrasts AD mice with wild type mouse at the timepoint of 6 months. The names 
of the vector are the genes and values are the p-value calculated from DESeq2.

```{r load_example}
library(LRcell)
data("example_gene_pvals")
head(example_gene_pvals, n=5)
```

Then, we applied LRcell analysis on this DEG list using the FC marker genes we just
acquired from the last section. Please note that the method for acquiring marker genes
and running LRcell should be the same.

```{r run_LRcell}
res <- LRcellCore(gene.p = example_gene_pvals,
           marker.g = FC_marker_genes,
           method = "LR", min.size = 5, 
           sig.cutoff = 0.05)
## curate cell types
res$cell_type <- unlist(lapply(strsplit(res$ID, '\\.'), '[', 2))
head(subset(res, select=-lead_genes))
```

Plot out the result using a function provided by LRcell:
```{r plot_LRcellCore, fig.width=10, fig.height=6, dpi=120} 
plot_manhattan_enrich(res, sig.cutoff = .05, label.topn = 5)
```

For further usage of LRcell, please check [LRcell Tutorial](https://github.com/marvinquiet/LRcell).

# sessionInfo()

```{r sessionInfo}
sessionInfo()
```
