---
title: "Description and Usage of MsBackendMassbank"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and Usage of MsBackendMassbank}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Spectra}
    %\VignetteDepends{Spectra,BiocStyle,RSQLite}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendMassbank")`<br />
**Authors**: `r packageDescription("MsBackendMassbank")[["Author"]] `<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(BiocStyle)
```

# Introduction

The `r Biocpkg("Spectra")` package provides a central infrastructure for the
handling of mass spectrometry (MS) data. The package supports interchangeable
use of different *backends* to import MS data from a variety of sources (such as
mzML files). The `r Biocpkg("MsBackendMassbank")` package allows import and
handling MS/MS spectrum data from
[Massbank](https://massbank.eu/MassBank/). This vignette illustrates the usage
of the *MsBackendMassbank* package to include MassBank data into MS data
analysis workflow with the *Spectra* package in R.

# Installation

The package can be installed with the *BiocManager* package. To
install *BiocManager* use `install.packages("BiocManager")` and, after that,
`BiocManager::install("MsBackendMassbank")` to install this package.

# Importing MS/MS data from MassBank files

MassBank is an open-source, community managed spectral library. All data is
available in the [MassBank GitHub](https://github.com/MassBank/MassBank-data)
page, where releases are provided (which are also shared through Zenodo, with
their own release-specific DOI). MassBank stores and shares data through
individual text files (one file per spectrum) in a specific MassBank
format. These files can be imported (as well as exported) with the
`MsBackendMassbank` class of the `r Biocpkg("MsBackendMassbank")` package.

In our example below we load the required libraries and define the (full) paths
to example MassBank files available in this package.

```{r load-libs}
library(Spectra)
library(MsBackendMassbank)
fls <- dir(system.file("extdata", package = "MsBackendMassbank"),
           full.names = TRUE, pattern = "txt$")
fls
```

MS data can be accessed and analyzed through `Spectra` objects. Below we create
a `Spectra` object with the data from these MassBank files. To this end we
provide the file names and specify to use a `MsBackendMassbank()` backend as
*source* to enable data import.

```{r import, warning = FALSE}
sps <- Spectra(fls,
               source = MsBackendMassbank(),
               backend = MsBackendDataFrame(),
               nonStop = TRUE)
```

With that we have now full access to all imported spectra variables (spectrum
metadata fields) that we list below.

```{r spectravars}
spectraVariables(sps)
```

We can for example access the *compound name* for each spectrum.

```{r}
sps$name
```

MassBank allows defining more than one name for a compound and the result is
thus returned as a `list` with all provided names and aliases per spectrum.

By default only some of the metadata fields available in the MassBank files are
imported. Through the `metaBlocks` parameter it is possible to enable also
import of additional blocks of metadata fields (which results however in a
slower data import). Below we use the `metaDataBlocks()` function to configure
the blocks to import. We select to import the `$AC` and `$MS` fields:

```{r metadata}
#' define the metadata blocks to import
mdb <- metaDataBlocks(ac = TRUE, ms = TRUE)

#' import the data
sps <- Spectra(fls,
               source = MsBackendMassbank(),
               metaBlock = mdb)
```

A larger number of spectra variables is now available:

```{r}
spectraVariables(sps)
```

For some of these, however, no information might be provided. To remove spectra
variables that have only missing values for **all** spectra, we can use the
`dropNaSpectraVariables()` function:

```{r}
sps <- dropNaSpectraVariables(sps)
spectraVariables(sps)
```

When importing a large number of MassBank files, setting `nonStop = TRUE`
prevents the call to stop whenever problematic MassBank files are
encountered.

# Accessing the MassBank MySQL database

An alternative to the import of the MassBank data from individual text files
(which can take a considerable amount of time) is to directly access the MS/MS
data in the MassBank MySQL database. For demonstration purposes we are using
here a tiny subset of the MassBank data which is stored as a SQLite database
within this package.

## Pre-requisites

At present it is not possible to directly connect to the main MassBank
*production* MySQL server, thus, to use the `MsBackendMassbankSql` backend it is
required to install the database locally. The MySQL database dump for each
MassBank release can be downloaded the MassBank GitHub repository (for most
releases). This dump could be imported to a local MySQL server.

## Direct access to the MassBank database

To use the `MsBackendMassbankSql` it is required to first connect to a
*MassBank* database. Below we show the R code which could be used for that - but
the actual settings (user name, password, database name, or host) will depend on
where and how the MassBank database was installed.

```{r mysql, eval = FALSE}
library(RMariaDB)
con <- dbConnect(MariaDB(), host = "localhost", user = "massbank",
                 dbname = "MassBank")
```

To illustrate the general functionality of this backend we use a tiny subset of
the MassBank (release 2020.10) which is provided as an small SQLite database
within this package. Below we connect to this database.

```{r sqlite}
library(RSQLite)
con <- dbConnect(SQLite(), system.file("sql", "minimassbank.sqlite",
                                       package = "MsBackendMassbank"))
```

We next *initialize* the `MsBackendMassbankSql` backend which supports direct
access to the MassBank in a SQL database and create a `Spectra` object from
that.

```{r}
mb <- Spectra(con, source = MsBackendMassbankSql())
mb
```

We can now use this `Spectra` object to access and use the MassBank data for our
analysis. Note that the `Spectra` object itself does not contain any data from
MassBank. Any data will be fetched on demand from the database backend.

To get a listing of all available annotations for each spectrum (the so-called
*spectra variables*) we can use the `spectraVariables()` function.

```{r}
spectraVariables(mb)
```

Through the `MsBackendMassbankSql` we can thus access spectra information as
well as its annotation.

We can access *core* spectra variables, such as the MS level with the
corresponding function `msLevel()`.

```{r}
head(msLevel(mb))
```

Spectra variables can also be accessed with `$` and the name of the
variable. Thus, MS levels can also be accessed with `$msLevel`:

```{r}
head(mb$msLevel)
```

In addition to spectra variables, we can also get the actual peaks (i.e. m/z and
intensity values) with the `mz()` and `intensity()` functions:

```{r}
mz(mb)
```

Note that not all spectra from the database were generated using the same
instrumentation. Below we list the number of spectra for each type of
instrument.

```{r}
table(mb$instrument_type)
```

We next subset the data to all spectra from ions generated by electro spray
ionization (ESI).

```{r}
mb <- mb[mb$ionization == "ESI"]
length(mb)
```

As a simple example to illustrate the `Spectra` functionality we next calculate
spectra similarity between one spectrum against all other spectra in the
database. To this end we use the `compareSpectra()` function with the normalized
dot product as similarity function and allowing 20 ppm difference in m/z between
matching peaks

```{r}
library(MsCoreUtils)
sims <- compareSpectra(mb[11], mb[-11], FUN = ndotproduct, ppm = 40)
max(sims)
```

We plot next a mirror plot for the two best matching spectra.

```{r}
plotSpectraMirror(mb[11], mb[(which.max(sims) + 1)], ppm = 40)
```

We can also retrieve the *compound* information for these two best matching
spectra. Note that this `compounds()` function works only with the
`MsBackendMassbankSql` backend as it retrieves the corresponding information
from the database's compound annotation table.

```{r}
mb_match <- mb[c(11, which.max(sims) + 1)]
compounds(mb_match)
```

Note that the `MsBackendMassbankSql` backend does not support parallel
processing because the database connection within the backend can not be shared
across parallel processes. Any function on a `Spectra` object that uses a
`MsBackendMassbankSql` will thus (silently) disable any parallel processing,
even if the user might have passed one along to the function using the `BPPARAM`
parameter. In general, the `backendBpparam()` function can be used on any
`Spectra` object to test whether its backend supports the provided parallel
processing setup (which might be helpful for developers).

# Session information

```{r}
sessionInfo()
```
