---
title: "LOB/LOD Estimation Workflow"
author: "Cyril Galitzine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LOB/LOD Estimation Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 


Load the required packages

```{r error=FALSE, warning=FALSE, message=FALSE}
library(MSstatsLOBD)
library(dplyr)
```

This Vignette provides an example workflow for how to use the package 
MSstatsLOBD.

## Installation

To install this package, start R (version "4.0") and enter:

``` {r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("MSstatsLOBD")
```


# 1 Example dataset

## 1 Introduction

We will estimate the LOB/LOD for a few peptides an assay available on the CPTAC 
(Clinical Proteomic Tumor Analysis Consortium) assay portal c.f. [@Thomas]. The 
dataset contains spike in data for 43 distinct peptides. For each peptide, 8 
distinct concentration spikes for 3 different replicates are measured. The 
Skyline files for the assay along with details about the experiment can be 
obtained from this webpage: https://assays.cancer.gov. The particular dataset 
examined here (called JHU_DChan_HZhang_ZZhang) can be found at <https://panoramaweb.org/labkey/project/CPTAC%20Assay%20Portal/JHU_DChan_HZhang_ZZhang/Serum_QExactive_GlycopeptideEnrichedPRM/begin.view?>.
It should be downloaded from the MSStats website 
<http://msstats.org/?smd_process_download=1&download_id=548>. 

The data is then exported in a csv file (${\tt calibration\_data\_raw.csv}$) from Skyline. This 
is done in Skyline by selecting File $\rightarrow$ Export $\rightarrow$ Report 
$\rightarrow$ QuaSAR input and then clicking Export. The csv file contains the 
measured peak area for each fragment of each light and heavy version of each peptide.
Depending on the format of the Skyline file and depending on whether standards 
were used, the particular outputs obtained in the csv file may vary. In this 
particular case the following variables are obtained in the output file 
${\tt calibration\_data\_raw.csv}$: ${\tt File\ Name,	Sample\ Name,	Replicate\ Name,	Protein\ Name,	Peptide\ Sequence,	Peptide\ Modified\ Sequence,}$	
${\tt Precursor\ Charge,	Product\ Charge,}$	${\tt Fragment\ Ion,	Average\ Measured\ Retention\ Time}$,
${\tt SampleGroup,	IS\ Spike,}$	${\tt Concentration,	 Replicate,light\ Area,	heavy\ Area}$. 
A number of variables are byproducts of the acquisition process and will not be 
considered for the following, i.e. 
${\tt File\ Name,	Sample\ Name,	Replicate\ Name, SampleGroup,	IS\ Spike}$.  
Variables that are important for the assay characterization are detailed below 
(others are assumed to be self explanatory):

 + ${\tt	Pepdide sequence}$ Name of the peptide sequence
 + ${\tt	Concentration}$ Value of the known spiked concentration in pmol.
 + ${\tt	Replicate}$ Number of the technical replicate
 + ${\tt	light\ Area}$ Peak area of the light (measured)
 + ${\tt	heavy\ Area}$ Peak area of the heavy (reference) peptide


## 2 Loading and Normalization of the data

### 2.1 Load the raw data file and check its content.

```{r}
head(raw_data)
```

### 2.2 Normalize Dataset

We normalize the intensity of the light peptides using that of the heavy 
peptides. This corrects any systematic errors that can occur during a run or 
across replicates. The calculation is greatly simplified by the use of the 
${\tt tidyr}$ and ${\tt dplyr}$ packages. The area from all the different 
peptide fragments is first summed then log transformed. The median intensity of 
the reference heavy peptides ${\tt medianlog2heavy}$ is calculated. Their 
intensities should ideally remain constant across runs since the spiked 
concentration of the heavy peptide is constant. The difference between the 
median for all the heavy peptide spikes is calculated. It is then used to 
correct (i.e. to normalize) the intensity of the light peptides 
${\tt log2light}$ to obtain the adjusted intensity ${\tt log2light\_norm}$. 
The intensity is finally converted back to original space.

```{r tidy=FALSE, warning=FALSE}
#Select variable that are need
df <- raw_data %>% select(Peptide.Sequence,Precursor.Charge, 
                          Product.Charge, Fragment.Ion,Concentration, 
                          Replicate, light.Area, heavy.Area, 
                          SampleGroup, File.Name)

#Convert factors to numeric and remove NA values:
df <- df %>% mutate(heavy.Area = as.numeric(heavy.Area)) %>% 
  filter(!is.na(heavy.Area))

#Sum area over all fragments
df2 <- df %>% group_by(Peptide.Sequence, Replicate, SampleGroup, 
                       Concentration, File.Name) %>%
  summarize(A_light = sum(light.Area), A_heavy = sum(heavy.Area))

#Convert to log scale
df2 <- df2 %>% mutate(log2light = log2(A_light), log2heavy = log2(A_heavy))

#Calculate median of heavy(reference) for a run
df3 <- df2 %>% group_by(Peptide.Sequence) %>% 
  summarize(medianlog2light = median(log2light), 
            medianlog2heavy= median(log2heavy))

#Modify light intensity so that the intensity of the heavy is constant (=median) across a run.
df4 <- left_join(df2,df3, by = "Peptide.Sequence") %>% 
  mutate(log2light_delta = medianlog2light - log2light) %>% 
  mutate(log2heavy_norm = log2heavy + log2light_delta, 
         log2light_norm = log2light + log2light_delta) %>% 
  mutate(A_heavy_norm = 2**log2heavy_norm, A_light_norm = 2**log2light_norm)

#Format the data for MSstats:
#Select the heavy area, concentration, peptide name and Replicate
df_out <- df4 %>% ungroup() %>% 
  select(A_heavy_norm, Concentration, Peptide.Sequence, Replicate)

#Change the names according to MSStats requirement:
df_out <- df_out %>% rename(INTENSITY = A_heavy_norm, 
                            CONCENTRATION = Concentration, 
                            NAME = Peptide.Sequence, REPLICATE = Replicate) 

# We choose NAME as the peptide sequence
head(df_out)
```


# 3 LOB/LOD definitions

## 3.1 Assay characterization procedure

\begin{figure}[ht] 
\centerline{\includegraphics[width=250pt]{LOD_example_3.pdf} }
\caption{Calculation of the LOB and LOD for peptide FLNDTMAVYEAK of the dataset}
\label{definition}
\end{figure}

In the following we estimate the LOB and LOD for individual peptides. The first 
step in the estimation is to fit a function to all the (Spiked Concentration, 
Measured Intensity) points. When the ${\tt nonlinear\_quantlim}$ function is 
used, the function that is fit automatically adapts to the data. For instance, 
when the data is linear, a straight line is used, while when a threshold (i.e. 
a leveling off of the measured intensity at low concentrations) an elbow like 
function is fit. The fit is called ${\tt MEAN}$ in the output of function 
${\tt nonlinear\_quantlim}$  as shown in Fig.1. Each value of ${\tt MEAN}$ is 
given for a particular ${\tt CONCENTRATION}$ value ${\tt CONCENTRATION}$ is 
thus a discretization of x--Spiked Concentration axis. The lower and upper 
bound of the 90% prediction interval of the fit are called ${\tt LOW}$  and 
${\tt UP}$ in the output of ${\tt nonlinear\_quantlim}$. They correspond 
respectively to the 5% and 95% percentile of predictions.

The second step in the procedure is to estimate the upper bound of the noise in 
the blank sample (blue dashed line in Fig. 1). It is found by assuming that 
blank sample measurements are normally distributed.

## 3.2 LOB/LOD definitions 

We define the LOB as the highest apparent concentration of a peptide expected 
when replicates of a blank sample containing no peptides are measured. This 
amounts to finding the concentration at the intersection of the fit (which 
represents the averaged measured intensity) with the 95% upper prediction bound 
of the noise.

The LOD is defined as the measured concentration value for which the probability
of falsely claiming the absence of a peptide in the sample is 0.05, given a 
probability 0.05 of falsely claiming its presence. Estimating the LOD thus 
amounts to finding the concentration at the intersection between the 5% 
percentile line of the prediction interval of the fit (i.e. the lower bound of 
the 90% prediction interval) and the 95% percentile line of the blank sample. 
At the LOB concentration, there is an 0.05 probability of false positive and a 
50% chance of false negative. At the LOD concentration there is 0.05 probability
of false negative and a false positive probability of 0.05 in accordance with 
its definition. By default, a probability of 0.05 for the LOB/LOD estimation is 
used but it can be changed, as detailed in the manual.

# 4 Estimation of the LOB/LOD for dataset

## 4.1 LOB/LOD estimation for a non-linear peptide

```{r warning=FALSE}

#Select peptide of interest:  LPPGLLANFTLLR
spikeindata <- df_out %>% filter(NAME == "LPPGLLANFTLLR")

#This contains the measured intensity for the peptide of interest
head(spikeindata)

#Call MSStatsLOD function:
quant_out <- nonlinear_quantlim(spikeindata)

head(quant_out)
```

After estimating LOB/LOD we can plot the results.

```{r error=FALSE, warning=FALSE, message=FALSE}
#plot results in the directory
plot_quantlim(spikeindata = spikeindata, quantlim_out = quant_out, address =  FALSE)
```

The threshold is captured by the fit at low concentrations. The ${\tt MEAN}$ of 
the output of the function is the red line (mean prediction) in the plots. 
${\tt LOW}$ is the orange line (5% percentile of predictions) while ${\tt UP}$ 
is the upper boundary of the red shaded area. The LOB is the concentration at 
the intersection of the fit and the estimate for the 95% upper bound of the 
noise (blue line). A more accurate "smoother" fit can be obtained by increasing 
the number of points ${\tt Npoints}$ used to discretize the concentration axis 
(see manual for ${\tt nonlinear\_quantlim}$).


The nonlinear MSStats function (${\tt nonlinear\_quantlim}$) works for all 
peptides (those with a linear response and those with a non-linear response). We
now examine a peptide with a linear behavior.


## 4.2 LOB/LOD estimation for a linear peptide

```{r warning=FALSE, comment=FALSE, warning=FALSE}
#Select peptide of interest:  FLNDTMAVYEAK
spikeindata2 <- df_out %>% filter(NAME == "FVGTPEVNQTTLYQR")

#This contains the measured intensity for the peptide of interest
head(spikeindata2)
```

```{r warning=FALSE}
#Call MSStats function:
quant_out2 <- nonlinear_quantlim(spikeindata2)

head(quant_out2)
```

```{r error=FALSE, warning=FALSE, message=FALSE}
#plot results in the directory: "/Users/cyrilg/Desktop/Workflow/Results"
#Change directory appropriately for your computer
plot_quantlim(spikeindata = spikeindata2, quantlim_out  = quant_out2, 
              address = FALSE)
```


The plots indicate that the fit is observed to be linear as the response is 
linear.

## 4.3 LOB/LOD linear estimation for a non-linear peptide

```{r warning=FALSE}

#Call MSStatsLOD function:
quant_out <- linear_quantlim(spikeindata)

head(quant_out)
```

After estimating LOB/LOD we can plot the results.

```{r error=FALSE, warning=FALSE, message=FALSE}
#plot results in the directory
plot_quantlim(spikeindata = spikeindata, quantlim_out = quant_out, address =  FALSE)
```

## 4.4 LOB/LOD linear estimation for a linear peptide

```{r warning=FALSE}

#Call MSStatsLOD function:
quant_out <- linear_quantlim(spikeindata2)

head(quant_out)
```

After estimating LOB/LOD we can plot the results.

```{r error=FALSE, warning=FALSE, message=FALSE}
#plot results in the directory
plot_quantlim(spikeindata = spikeindata2, quantlim_out = quant_out, address =  FALSE)
```

## REFERENCES

C. Galitzine et al. “Nonlinear regression improves accuracy of characterization 
of multiplexed mass spectrometric assays.” Molecular & Cellular Proteomics 
(2018), doi:10.1074/mcp.RA117.000322

